---
title: "Mental Health Among Incarcerated Women"
author: "Abbie Tolon"
date: "January 29, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
###Set-up

First, I installed all of the necessary packages and loaded them to my library. Next, I set the working directory, where the original SAS files of the raw data were stored. The raw SAS files were then converted to Excel documents and saved to the working directory. Within the Excel document, unnecessary variables were removed, keeping only the variables of interest. Moreover, only those who identified as "female" were kept for the analysis. This left 30 variables of interest for analysis and a total of 1993 observations (originally, there were 6982 total observations and 2487 variables). Finally, the Excel documents were uploaded in R to prepare for further cleaning and analysis. 

```{r, echo=FALSE, results='hide'}
# installing/loading the latest installr package:
#install.packages("installr"); library(installr) # install+load installr

#updateR() # updating R.

#ONLY ONLY ONLY for first loading of a package, you must install

#load packages
library(tidyverse) # for ggplot2,tibble,tidyr,readr,purrr,dplyr,stringr,forcats
library(haven) # for reading Stata files
library(labelled) # for working with variable labels
library(magrittr) # for extra %$% pipe operator
library(readxl) # for reading the excel file
library(writexl) # for saving something as an Excel file
library(sas7bdat) # for opening a SAS file
library(dplyr) # gives rename function

##Set Working Directory --------------------------------------------------------------------------------
#setwd("C:/Users/artol/Documents/Fl19/CapstoneR/Data")

##Name loaded file as "jail"
#Jail1 <- read_sas("test1.sas7bdat")
#Jail2 <- read_sas("test2.sas7bdat")

##Inspect the data
#view(Jail1)
#view(Jail2)

##save the r-file as an Excel file, so that we can manipulate data in spreadsheet form
#write_xlsx(x = Jail1, path = "jail1.xlsx", col_names = TRUE)
#write_xlsx(x = Jail2, path = "jail2.xlsx", col_names = TRUE)

#Retrieve Data from Github -----------------------------------------------------------------------------
jail <-read.csv("https://raw.githubusercontent.com/artolon/Capstone/master/CapstoneR/Data/jail1keydata.csv?token=AOF354NV2SKFHAYG4PXUKEC6GORC4")

#View quick scan of the data
summary(jail)
```
### Data Cleaning

First, the dichotomous dependent variable was created (mental_illness). This was done by aggregating every variable related to mental illness. Such variables included: Depression, bipolar, anxiety, psychotic disorder, personality disorder, PTSD, and "other." If a person marked "yes" for any of those variables, "mental_illness" was coded as 1. If all variables were marked as "no," then "mental_illness" was also marked as "no." All missing variables were coded as "99." 

Next, the effect modifier, "have a disability," was recoded so that 1=yes, 0=no, and 99=missing. 

The following independent variables were then recoded, so that 1=yes, 0=no, and 99=missing: "parents/guardians abused alcohol/drugs," "parents/step-parents served time," "ever been shot at," "physically abused prior to current admission," and "ever been sexually assaulted/molested." These variables were selected, because they are examples of Adverse Childhood Experiences (ACEs) and past trauma. 

Finally, confounders (race, education, and income) were all recoded. Race variables were recoded so that if a person identified with a particular race ("yes"), the variable would be marked as "1." Missing variables were coded to 99. The education variable was recoded into fewer categories. Specifically, "No school through 11th grade," "12th grade," "freshman through junior year of college," and "senior in college or higher." All other values were coded as "missing=99." One limitation to this variable, is that it does not report whether a person obtained a degree. Therefore, it must be interpreted as "years of education," rather than "degree obtained." Finally, the income variable was also recoded into fewer categories. This variable describes a person's monthly income in the most prior to arrest. The categories were recoded to the following: 0="No income"; 1="$1-$499"; 2="$500-$999"; 3="$1000-$1999"; 4="$2000-$4999"; 5="$5000 or more";and 99="missing."

```{r, echo=FALSE, results='hide'}
#***Dependent Variable***
#First, create a dichotomous variable called "mental_illness"; create this by aggregating
#all the mental health variables. If ANY=1, the variable is marked 1. Otherwise, it is 0. If missing, 
#the variable will = 99

#Check class of variable before recoding 
class(jail$V2022Depression) #is integer

#Change class to character for every MH variable for recording (we will later change back to numeric)
jail$V2022Depression <- as.character(jail$V2022Depression)
jail$V2023Bipolar <- as.character(jail$V2023Bipolar)
jail$V2026OtherAnxiety <- as.character(jail$V2026OtherAnxiety)
jail$V2024PsychoticDisorder <- as.character(jail$V2024PsychoticDisorder)
jail$V2027PersonalityDis <- as.character(jail$V2027PersonalityDis)
jail$V2025PTSD <- as.character(jail$V2025PTSD)
jail$V2028Other <- as.character(jail$V2028Other)

#Check class again to make sure everything coded correctly 
class(jail$V2022Depression) #is character
class(jail$V2023Bipolar) #is character
class(jail$V2026OtherAnxiety) #is character
class(jail$V2024PsychoticDisorder) #is character
class(jail$V2027PersonalityDis) #is character
class(jail$V2025PTSD) #is character
class(jail$V2028Other) #is character

#Now, create a new variable called "mental_illness," where mental_illness=1 if all related variables=1, and where
#mental_illness=0, if all related variables =2; otherwise, mental_illness=99 (missing)

names(jail) #to get the variable/column names, so that I don't have to keep going back and forth

#Create the new variable
jail2 <- jail %>%
  mutate(mental_illness=ifelse(V2022Depression == 1|V2023Bipolar==1|V2026OtherAnxiety==1|
                                 V2024PsychoticDisorder==1|V2027PersonalityDis==1|V2025PTSD==1|
                                 V2028Other==1, 1, ifelse(V2022Depression ==2|V2023Bipolar==2|V2026OtherAnxiety==2|
                                                  V2024PsychoticDisorder==2|V2027PersonalityDis==2|V2025PTSD==2|
                                                            V2028Other==2, 0, 99)))

#View variable to see if coding was completed correctly 
view(jail2$mental_illness)

#create crosstabs to inspect some of the variables to ensure the coding was completed successfully
table(jail2$mental_illness,jail2$V2022Depression)
xtabs(~jail2$mental_illness + jail2$V2023Bipolar + jail2$V2024PsychoticDisorder)
```

```{r, echo=FALSE, results='hide'}
#***Effect Modifier***

#Check class of variable
class(jail2$V2054Disability) #is integer

#Change to character variable 
jail2$V2054Disability <- as.character(jail2$V2054Disability)

#Check class
class(jail2$V2054Disability) #is character

#We are going to recode the disability variable, so that 1=yes, 0=no, and 99=missing
jail3 <- jail2 %>%
  mutate(Disability=ifelse(V2054Disability==1, 1, ifelse(V2054Disability==2, 0, 99)))

#Check the data to make sure it was done correctly
table(jail3$Disability, jail3$V2054Disability)

#View a quick scan of the variable, by itself
view(jail3$Disability)
```

```{r, echo=FALSE, results='hide'}
#***Independent Variables***

#The independent variables are as follows: 
#"parents/guardians abused alcohol/drugs," "parents/step-parents served time," "ever been shot at," "physically abused prior to current admission," and "ever been sexually assaulted/molested." 

#Check class of variable before recoding 
class(jail3$V1509ParentsAbusedAlcDrug) #is integer

#Change class to character for every independent variable for recording (we will later change back to numeric)
jail3$V1509ParentsAbusedAlcDrug <- as.character(jail3$V1509ParentsAbusedAlcDrug)
jail3$V1512ParentsServedTime <- as.character(jail3$V1512ParentsServedTime)
jail3$V1537EverShot <- as.character(jail3$V1537EverShot)
jail3$V1577PhysicalAbuse <- as.character(jail3$V1577PhysicalAbuse)
jail3$V1541SexualAssault <- as.character(jail3$V1541SexualAssault)

#Check class again to make sure everything coded correctly 
class(jail3$V1509ParentsAbusedAlcDrug) #is character
class(jail3$V1512ParentsServedTime) #is character
class(jail3$V1537EverShot) #is character
class(jail3$V1577PhysicalAbuse) #is character
class(jail3$V1541SexualAssault) #is character

#recode "parents/guardians abused alcohol/drugs" - V1509ParentsAbusedAlcDrug
    #Currently, 1=yes, 2=no, 7=don't know, 8=refused, 9=blank
jail3 <- jail3 %>%
  mutate(ParentsAbusedAlcDrug=ifelse(V1509ParentsAbusedAlcDrug==1, 1, ifelse(V1509ParentsAbusedAlcDrug==2, 0, 99)))

#recode "parents/step-parents served time" - V1512ParentsServedTime
    #Currently, 1=yes, 2=no, 7=don't know, 8=refused, 9=blank
jail3 <- jail3 %>%
  mutate(ParentsServedTime=ifelse(V1512ParentsServedTime==1, 1, ifelse(V1512ParentsServedTime==2, 0, 99)))

#recode "ever been shot at" - V1537EverShot
    #Currently, 1=yes, 2=no, 7=don't know, 8=refused, 9=blank
jail3 <- jail3 %>%
  mutate(EverShot=ifelse(V1537EverShot==1, 1, ifelse(V1537EverShot==2, 0, 99)))

#recode "physically abused prior to current admission" - V1577PhysicalAbuse
    #Currently, 1=yes, 2=no, 7=don't know, 8=refused, 9=blank
jail3 <- jail3 %>%
  mutate(PhysicalAbuse=ifelse(V1577PhysicalAbuse==1, 1, ifelse(V1577PhysicalAbuse==2, 0, 99)))

#recode "ever been sexually assaulted/molested" - V1541SexualAssault
    #Currently, 1=yes, 2=no, 7=don't know, 8=refused, 9=blank
jail3 <- jail3 %>%
  mutate(SexualAssault=ifelse(V1541SexualAssault==1, 1, ifelse(V1541SexualAssault==2, 0, 99)))

#Check to confirm variables were recoded 
table(jail3$ParentsAbusedAlcDrug, jail3$V1509ParentsAbusedAlcDrug)
table(jail3$ParentsServedTime, jail3$V1512ParentsServedTime)
table(jail3$EverShot, jail3$V1537EverShot)
table(jail3$PhysicalAbuse, jail3$V1577PhysicalAbuse)
table(jail3$SexualAssault, jail3$V1541SexualAssault) #tables look good
```
```{r, echo=FALSE, results='hide'}
#***Confounding Variables***
names(jail3)

#Change class to character for every confounding variable for recording (we will later change back to numeric)
jail3$V28White <- as.character(jail3$V28White)
jail3$V29Black <- as.character(jail3$V29Black)
jail3$V30AmericanIndianAlaskaNative <- as.character(jail3$V30AmericanIndianAlaskaNative)
jail3$V31Asian <- as.character(jail3$V31Asian)
jail3$V32NativeHawaiinPacificIslander <- as.character(jail3$V32NativeHawaiinPacificIslander)
jail3$V33OtherRace <- as.character(jail3$V33OtherRace)
jail3$V34UnknownRace <- as.character(jail3$V34UnknownRace)
jail3$V44Black <- as.character(jail3$V44Black)
jail3$V1391Education <- as.character(jail3$V1391Education)
jail3$V150Income <- as.character(jail3$V150Income)

#Check class again to make sure everything coded correctly 
class(jail3$V28White) #is character
class(jail3$V29Black) #is character
class(jail3$V30AmericanIndianAlaskaNative) #is character
class(jail3$V31Asian) #is character
class(jail3$V32NativeHawaiinPacificIslander) #is character
class(jail3$V33OtherRace) #is character
class(jail3$V34UnknownRace) #is character
class(jail3$V44Black) #is character
class(jail3$V1391Education) #is character
class(jail3$V150Income) #is character

#------------------------------------------------------------------------------------------------------------------
#Recode Race variables: V28White, V29Black, V30AmericanIndianAlaskaNative, V31Asian, V32NativeHawaiinPacificIslander, V33OtherRace, V34UnknownRace, V44Black
    #Currently...
    #V28White (1=yes; 99=Blank)
    #V29Black (2=yes; 99=Blank)
    #V30AmericanIndianAlaskaNative (3=yes; 99=Blank)
    #V31Asian (4=yes; 99=Blank)
    #V32NativeHawaiinPacificIslander (5=yes; 99=Blank)
    #V33OtherRace (6=yes; 99=Blank)
    #V34UnknownRace (7=yes; 8=refused; 9=Blank)

#First, recode the race variables, so that 1=yes for all of the options 
jail3 <- jail3 %>%
  mutate(White=ifelse(V28White==1, 1, 99))

jail3 <- jail3 %>%
  mutate(Black=ifelse(V29Black==2, 1, 99))

jail3 <- jail3 %>%
  mutate(AmericanIndianAlaskaNative=ifelse(V30AmericanIndianAlaskaNative==3, 1, 99))

jail3 <- jail3 %>%
  mutate(Asian=ifelse(V31Asian==4, 1, 99))

jail3 <- jail3 %>%
  mutate(NativeHawaiinPacificIslander=ifelse(V32NativeHawaiinPacificIslander==5, 1, 99))

jail3 <- jail3 %>%
  mutate(OtherRace=ifelse(V33OtherRace==6, 1, 99))

jail3 <- jail3 %>%
  mutate(UnknownRace=ifelse(V34UnknownRace==7, 1, 99))

table(jail3$White, jail3$V28White) #1103 people
table(jail3$Black, jail3$V29Black) #787 people
table(jail3$AmericanIndianAlaskaNative, jail3$V30AmericanIndianAlaskaNative) #110 people
table(jail3$Asian, jail3$V31Asian) #11 people
table(jail3$NativeHawaiinPacificIslander, jail3$V32NativeHawaiinPacificIslander) #12 people
table(jail3$OtherRace, jail3$V33OtherRace) #77 people
table(jail3$UnknownRace, jail3$V34UnknownRace) #0 people

#Now, recode the dichotomous variable, called V44Black
    #V44Black (1=Black; 2=NonBlack); recode so that 1=yes and 0=no
jail3 <- jail3 %>%
  mutate(Black2=ifelse(V44Black==1, 1, ifelse(V44Black==2, 0, 99)))

#Check variable to confrim it was recoded correctly 
table(jail3$Black2, jail3$V44Black) #recode worked; 788 marked "yes", which is 1 off from V29Black...

#-----------------------------------------------------------------------------------------------------------------
#Recode V1391Education (00=Never attended or attended kindergarten only; 01=first; 02=second; 03=third; 04=fourth; 05=fifth; 06=sixth; 07=seventh; 08=eighth; 09=ninth; 10=tenth; 11=eleventh; 12=twelfth; 13=freshman; 14=sophomore; 15=junior; 16=senior; 17=one year graduate school; 18=two or more years of graduate school; 97=don't know; 98=refused; 99=blank)

#I will recode this variables so that 00-11=1; 12=2; 13-15=3; 16-18=4 (this variable has issues because I don't know whether someone as a degree)
jail3 <- jail3 %>%
  mutate(Education=ifelse(V1391Education==00|V1391Education==01|V1391Education==02|V1391Education==03|V1391Education==04|V1391Education==05|V1391Education==06|V1391Education==07|V1391Education==08|V1391Education==09|V1391Education==10|V1391Education==11, 1, ifelse(V1391Education==12, 2, ifelse(V1391Education==13|V1391Education==14|V1391Education==15, 3, ifelse(V1391Education==16|V1391Education==17|V1391Education==18, 4, 99)))))

#Check variable to confirm it was recoded correctly
table(jail3$Education, jail3$V1391Education) #everything looks good!

#------------------------------------------------------------------------------------------------------------------
#Recode V1501Income (described as MONTHLY income in the month prior to arrest)
    #currently (00=no income; 01=$1-99; 02=100-199; 03=200-299; 04=300-399; 05=400-499; 06=500-599; 07=600-799; 08=800-999; 09=1000-1199; 10=1200-1499; 11=1500-1999; 12=2000-2499; 13=2500-4999; 14=5000 or more; 97=don't know; 98=refused; 99=blank)

#recoding so that 0=no income; 1=$1-499; 2=$500-999; 3=$1000-1999; 4=$2000-4999; 5=$5000+; 99=missing
jail3 <- jail3 %>%
  mutate(Income=ifelse(V150Income==00, 0, ifelse(V150Income==01|V150Income==02|V150Income==03|V150Income==04|V150Income==05, 1, ifelse(V150Income==06|V150Income==07|V150Income==08, 2, ifelse(V150Income==09|V150Income==10|V150Income==11, 3, ifelse(V150Income==12|V150Income==13, 4, ifelse(V150Income==14, 5, 99)))))))

#Check variable to confirm it was recoded correctly 
table(jail3$Income, jail3$V150Income) #table looks good! 

#------------------------------------------------------------------------------------------------------------------
#At the end of recoding...change ALL columns to numeric!!
jail3$White <- as.numeric(jail3$White)
jail3$Black <- as.numeric(jail3$Black)
jail3$AmericanIndianAlaskaNative <- as.numeric(jail3$AmericanIndianAlaskaNative)
jail3$Asian <- as.numeric(jail3$Asian)
jail3$NativeHawaiinPacificIslander <- as.numeric(jail3$NativeHawaiinPacificIslander)
jail3$OtherRace <- as.numeric(jail3$OtherRace)
jail3$UnknownRace <- as.numeric(jail3$UnknownRace)
jail3$Black2 <- as.numeric(jail3$Black2)
jail3$Education <- as.numeric(jail3$Education)
jail3$Income <- as.numeric(jail3$Income)

#Check class of the variables
class(jail3$White) #is numeric 
class(jail3$Black) #is numeric 
class(jail3$AmericanIndianAlaskaNative) #is numeric 
class(jail3$Asian) #is numeric 
class(jail3$NativeHawaiinPacificIslander) #is numeric 
class(jail3$OtherRace) #is numeric 
class(jail3$UnknownRace) #is numeric 
class(jail3$Black2) #is numeric 
class(jail3$Education) #is numeric 
class(jail3$Income) #is numeric 
```
###Explore Variables

Now that all applicable variables have been recoded, we are going to explore and summarize the data. First, we will analyze demographic variables, such as age, race, education, and income. 

```{r, echo=FALSE, results='hide'}
names(jail3)
#Age variable (V13Age)--------------------------------------------------------------------------------------------

#check class of variable
class(jail3$V13Age) #is integer

#change to numeric 
jail3$V13Age <- as.numeric(jail3$V13Age)

#Examine the quantiles 
quantile(jail3$V13Age, na.rm = TRUE) #median age = 33; range of 14 to 72

#Find avaerage age 
mean(jail3$V13Age, na.rm = TRUE) #average age = 33.21

#count missing variables 
sum(is.na(jail3$V13Age)) #only one value is missing 

#create a new variables that recodes age into a categorical variable 
jail3 <- jail3 %>%
  mutate(AgeCat=ifelse(V13Age<=17, 1, ifelse(V13Age>17 & V13Age<=34, 2, ifelse(
    V13Age>34 & V13Age <=50, 3, ifelse(V13Age>50 & V13Age<=99, 4, 99)))))

#check new variable
table(jail3$AgeCat, jail3$V13Age) #table looks good!

#Look at the breakdown for Age cat
sum(jail3$AgeCat == 1, na.rm = TRUE) #29 people are under 18
sum(jail3$AgeCat == 2, na.rm = TRUE) #1071 people between 18 and 34
sum(jail3$AgeCat == 3, na.rm = TRUE) #826 people between 35 and 50
sum(jail3$AgeCat == 4, na.rm = TRUE) #66 people are 51 and older

(29/1993)*100 # 1.46% are under 18
(1071/1993)*100 #53.74% are between 18 and 34
(826/1993)*100 #41.45% are between 35 and 50
(66/1993)*100 #3.31% are 51 and older 

#Race variables----------------------------------------------------------------------------------------------------

#Check class of the variables
sum(jail3$White==1, na.rm = TRUE) #1103 people are White
sum(jail3$Black==1, na.rm = TRUE) #787 people are Black
sum(jail3$AmericanIndianAlaskaNative==1, na.rm = TRUE) #110 people are American Indian/Alaska Native
sum(jail3$Asian==1, na.rm = TRUE) #11 people are Asian 
sum(jail3$NativeHawaiinPacificIslander==1, na.rm = TRUE) #12 people are Native Hawaiin/Pacific Islander 
sum(jail3$OtherRace==1, na.rm = TRUE) #77 people are an "other" race
sum(jail3$UnknownRace==1, na.rm = TRUE) #0 people are "unknown"
sum(jail3$Black2==1, na.rm = TRUE) #The dichotomous variable has 788 people marked as identifying as "Black"

(1103/1993)*100 #55.34% identified as White
(787/1993)*100 #39.49% identified as Black
(110/1993)*100 #5.52% identified as American Indian/Alaska Native
(11/1993)*100 #0.55% identified as Asian
(12/1993)*100 #0.60% identified as Native Hawaiin/Pacific Islander
(77/1993)*100 #3.86% identified as "other"
(0/1993)*100 #0% of people marked "unknown" for race

#Dichotmous variable for "Black" vs. "Nonblack" 
(788/1993)*100 #39.54 identified as "Black"
100-((788/1993)*100) #60.46 identified as "Nonblack"

#Education variable------------------------------------------------------------------------------------------------
sum(jail3$Education==1, na.rm = TRUE) #988 people completed 11th grade or lower
sum(jail3$Education==2, na.rm = TRUE) #594 people compelted 12th grade
sum(jail3$Education==3, na.rm = TRUE) #327 people completed their junior year of college
sum(jail3$Education==4, na.rm = TRUE) #62 people completed their senior year of college or more

(988/1993)*100 #49.57% completed 11th grade or lower
(594/1993)*100 #29.80% people compelted 12th grade
(327/1993)*100 #16.41% people completed their junior year of college
(62/1993)*100 #3.11% people completed their senior year of college or more

#Income variable---------------------------------------------------------------------------------------------------
sum(jail3$Income==0, na.rm = TRUE) #127 people had no income the month prior to arrest
sum(jail3$Income==1, na.rm = TRUE) #453 people earned $1-$499 the month prior to arrest
sum(jail3$Income==2, na.rm = TRUE) #395 people earned $500-$999 the month prior to arrest
sum(jail3$Income==3, na.rm = TRUE) #322 people earned $1000-$1999 the month prior to arrest
sum(jail3$Income==4, na.rm = TRUE) #144 people earned $2000-$4999 the month prior to arrest
sum(jail3$Income==5, na.rm = TRUE) #45 people earned $5000 or more the month prior to arrest

(127/1993)*100 #6.37% people had no income the month prior to arrest
(453/1993)*100 #22.73% people earned $1-$499 the month prior to arrest
(395/1993)*100 #19.82% people earned $500-$999 the month prior to arrest
(322/1993)*100 #16.16% people earned $1000-$1999 the month prior to arrest
(144/1993)*100 #7.23% people earned $2000-$4999 the month prior to arrest
(45/1993)*100 #2.26% people earned $5000 or more the month prior to arrest
```

